---
title: "Overview"
author:
- name: William Hutchison
  affiliation: WEHI - Walter and Eliza Hall Institute of Medical Research
  email: hutchison.w@wehi.edu.au
- name: Stefano Mangiola
  affiliation: WEHI - Walter and Eliza Hall Institute of Medical Research
package: tidySpatialExperiment
output:
  BiocStyle::html_document
abstract: |
  A brief overview of the tidySpatialExperiment package - demonstrating the SpatialExperiment-tibble abstraction, compatibility with the *tidyverse* ecosystem, compatibility with the *tidyomics* ecosystem and a few helpful utility functions.
vignette: |
  %\VignetteIndexEntry{Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

tidySpatialExperiment provides a bridge between the [SpatialExperiment](https://github.com/drighelli/SpatialExperiment) [@righelli2022spatialexperiment] package and the [*tidyverse*](https://www.tidyverse.org) [@wickham2019welcome] ecosystem. It creates an invisible layer that allows you to interact with a SpatialExperiment object as if it were a tibble; enabling the use of functions from [dplyr](https://github.com/tidyverse/dplyr), [tidyr](https://github.com/tidyverse/tidyr), [ggplot2](https://github.com/tidyverse/ggplot2) and [plotly](https://github.com/plotly/plotly.R). But, underneath, your data remains a SpatialExperiment object.

tidySpatialExperiment also provides three additional utility functions.

## Functions and utilities

Package | Functions available
------------ | -------------
`SpatialExperiment` | All
`dplyr` | `arrange`,`bind_rows`, `bind_cols`, `distinct`, `filter`, `group_by`, `summarise`,  `select`, `mutate`, `rename`, `left_join`, `right_join`, `inner_join`, `slice`, `sample_n`, `sample_frac`, `count`, `add_count`
`tidyr` | `nest`, `unnest`, `unite`, `separate`, `extract`, `pivot_longer`
`ggplot2` | `ggplot`
`plotly` | `plot_ly`

Utility | Description
------------ | -------------
`as_tibble` | Convert cell-wise information to a `tbl_df`
`join_features` | Combine cell-wise and feature-wise information into a `tbl_df`
`aggregate_cells` | Aggregate cell-feature abundance into a pseudobulk `SummarizedExperiment`

## SpatialExperiment-tibble abstraction

A SpatialExperiment object represents observations (cells) as columns and variables (features) as rows, as is the Bioconductor convention. Additional information about the cells is accessed through the `reducedDims`, `colData` and `spatialCoords` functions. 

tidySpatialExperiment provides a SpatialExperiment-tibble abstraction, representing cells as rows and features as columns, as is the *tidyverse* convention. `colData` and `spatialCoords` are appended as columns to the same abstraction, allowing easy interaction with this additional data. 

## Installation

You can install the development version of tidySpatialExperiment from GitHub with:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("william-hutchison/tidySpatialExperiment")
```

## Load data

Here, we load and view an example SpatialExperiment object. The output we see is of the SpatialExperiment-tibble abstraction.

```{r, message=FALSE, results=FALSE}
# Load example SpatialExperiment object
library(tidySpatialExperiment)
example(read10xVisium)
```

```{r, echo=FALSE}
# Load chromote for knitting
# library(chromote)
```

## View data

```{r}
# View the SpatialExperiment-tibble abstraction
spe
```

However, our data maintains its status as a SpatialExperiment object. Therefore, we have access to all SpatialExperiment functions. 

```{r}
spe |>
  colData() |>
  head()

spe |> 
  spatialCoords() |>
  head()

spe |>
  imgData()
```

# Integration with the *tidyverse* ecosystem
## Manipulate with dplyr

Most functions from dplyr are available for use with the SpatialExperiment-tibble abstraction. For example, `filter` can be used to select cells by a variable of interest. 

```{r}
spe |>
  filter(array_col < 5)

```

And `mutate` can be used to add new variables, or modify the value of an existing variable.
```{r}
spe |>
  mutate(in_region = c(in_tissue & array_row < 10))
```

## Tidy with tidyr

Most functions from tidyr are also available. Here, `nest` is used to group the data by `sample_id`, and `unnest` is used to ungroup the data.

```{r}
# Nest the SpatialExperiment object by sample_id
spe_nested <-
  spe |> 
  nest(data = -sample_id)

# View the nested SpatialExperiment object
spe_nested

# Unnest the nested SpatialExperiment objects
spe_nested |>
  unnest(data)
```

## Plot with ggplot2

The `ggplot` function can be used to create a plot from a SpatialExperiment object. This example also demonstrates how tidy operations can be combined to build up more complex analysis. It should be noted that helper functions such `aes` are not included and should be imported from ggplot2. 

```{r}
spe |>
  filter(sample_id == "section1" & in_tissue) |>
  
  # Add a column with the sum of feature counts per cell
  mutate(count_sum = purrr::map_int(.cell, ~
    spe[, .x] |> 
      counts() |> 
      sum()
    )) |>
  
  # Plot with tidySpatialExperiment and ggplot2
  ggplot(ggplot2::aes(x = reorder(.cell, count_sum), y = count_sum)) +
  ggplot2::geom_point() +
  ggplot2::coord_flip()
```

## Plot with plotly
The `plot_ly` function can also be used to create a plot from a SpatialExperiment object.

```{r, fig.keep="first", message=FALSE, warning=FALSE}
spe |>
  filter(sample_id == "section1") |>
  plot_ly(
    x = ~ array_col, 
    y = ~ array_row, 
    color = ~ in_tissue, 
    type = "scatter"
  )
```

# Integration with the *tidyomics* ecosystem 

## Interactively select cells with tidygate

Different packages from the *tidyomics* ecosystem are easy to use together. When run in the R Console, the following code will lunch tidygate to interactively gate cells based on their array location.

```{r, eval=FALSE}
spe_regions <-
  spe |> 
  filter(sample_id == "section1") |>
  mutate(region = tidygate::gate_chr(array_col, array_row))
```

```{r, echo=FALSE}
# Manually set gate information to match demonstration
spe_regions <-
  spe |> 
  filter(sample_id == "section1") |>
  mutate(region = ifelse(
    array_row < 48 &
    array_row > 20 &
    array_col < 80 &
    array_col > 60, 
    1, 0))
```

The gated cells can then be divided into pseudobulks within a SummarizedExperiment object using tidySpatialExperiment's `aggregate_cells` utility function. 

```{r}
spe_regions_aggregated <-
  spe_regions |>
  aggregate_cells(region)
```

# Important considerations

## Read-only columns

Removing the `.cell` column will return a tibble. This is consistent with the behaviour in other *tidyomics* packages.
```{r}
spe |>
  select(-.cell) |>
  head()
```

The sample_id column cannot be removed with *tidyverse* functions, and can only be modified if the changes are accepted by SpatialExperiment's `colData` function.

```{r, error=TRUE}
# sample_id is not removed, despite the user's request
spe |>
  select(-sample_id)

# This change maintains separation of sample_ids and is permitted
spe |> 
  mutate(sample_id = stringr::str_c(sample_id, "_modified")) |>
  head()

# This change does not maintain separation of sample_ids and produces an error
spe |>
  mutate(sample_id = "new_sample")
```

The `pxl_col_in_fullres` and `px_row_in_fullres` columns cannot be removed or modified with *tidyverse* functions. This is consistent with the behaviour of dimension reduction data in other *tidyomics* packages.

```{r, error=TRUE}
# Attempting to remove pxl_col_in_fullres produces an error
spe |>
  select(-pxl_col_in_fullres)

# Attempting to modify pxl_col_in_fullres produces an error
spe |> 
  mutate(pxl_col_in_fullres)
```

# Session info
```{r sessionInfo, echo=FALSE}
sessionInfo()
```